Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens

### 2014-01-28 limits/ms.tex
one word change.
Justin added text to support new Appendix Figures; deleted remaining highlights.

\title[Lens Recovery with \Glass]{Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens}
\author[J.~P.~Coles~et~al.]{%
Jonathan P.~Coles,$^{1,3}$\thanks{jonathan@exascale-computing.eu}%
\newauthor%
Justin I.~Read$^2$,%
\newauthor%
and Prasenjit Saha$^3$%
%
\\
$^1$Exascale Research Computing Lab, Universit\'e de Versailles Saint-Quentin-en-Yvelines,\\$\quad$55 Avenue des Etats Unis, 78000, Versailles, France\\
$^2$\hilight{Department of Physics, University of Surrey, Guildford, Surrey, GU2 7XH, United Kingdom} \\
$^3$\hilight{Institute of Physics, University of Zurich, 190 Winterthurerstrasse, 8057, Zurich, Switzerland}
}

@@

The key scientific and technical improvements are:
%
\begin{enumerate}
  \setcounter{enumi}{0}
  \item A modular framework that allows new priors to be added and modified easily.
\end{enumerate}
%
Each prior is a simple function that adds linear constraints that operate on either
a single lens object or the entire ensemble of objects. \Glass\ comes with a number
of useful priors, the default ones will be described later, but a user can
write their own directly in the input file, or by modifying the source code.
%
\begin{enumerate}
  \setcounter{enumi}{1}
  \item The basis functions approximating a model can be changed. 
\end{enumerate}
%
\Glass\ currently describes the lens mass as a collection of pixels, but the code
has been designed to support alternative methods. In particular, there are future
plans to develop a module using Bessel functions. This will require a new set of 
priors that operate on these functions.
%
\begin{enumerate}
  \setcounter{enumi}{2}
  \item Non-linear constraints can be imposed in an automated post-processing step. 
\end{enumerate}
%
Once \Glass\ has generated an ensemble of models given the linear constraints, any number
of post processing functions can be applied. Not only can these functions be used to
derive new quantities from the mass models, they can also be used as a filter to 
accept or reject a model based on some non-linear constraint. The plotting functions
within \Glass\ will correctly display models that have been accepted or rejected.
%
\begin{enumerate}
  \setcounter{enumi}{3}
  \item The central region can have a higher resolution to capture steep models. 
\end{enumerate}
%
With the default basis set of pixels, the mass distribution of the lens is
described by a uniform grid. However, in the central region of a lensing galaxy
where the mass profile may rise steeply, the center pixel uses a higher
resolution. This allows the density to increase smoothly but still allow for
a large degree of freedom within the inner region without allowing the density
to be arbitrarily high. 
%
\begin{enumerate}
  \setcounter{enumi}{4}
  \item Stellar density can be used as an additional constraint.  
\end{enumerate}
%
The mass in inner regions of galaxies is often dominated by the stellar component
which one can estimate using standard mass-to-light models. This data can be added
to the potential as described later in \secref{stellar mass}. By using the stellar
mass one can place a lower bound on the mass and help constrain the inner most
mass profile.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item Point or extended mass objects can be placed in the field.
\end{enumerate}
%
\PixeLens\ only allows for a shear term to be added to the potential (as shown
later in \eqnref{shear}) to account for mass external to the modelled region. This
is useful to capture the gross effects of a distant neighbour. \Glass\ also
has a shear term but additionally allows for analytic potentials to
be included. This can be used to model substructure or multiple neighbors close
to the main lens. The substructure may have only a small effect if the lens is
a single galaxy, but if the lens is a group then a potential can be added for
each of the galaxies. A few standard functions are already included in \Glass\
including those for a point mass, a power law distribution, or an isothermal (a
particular case of the power law).
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item A new uniform sampling algorithm for high dimensional spaces.
\end{enumerate}
%
At the heart of \Glass\ lies a new algorithm for sampling the high dimensional
linear space that represents the modelling solution space. This algorithm was
described and tested in \cite{2012MNRAS.425.3077L}; it is 
multi-threaded allowing it to run efficiently on many-cored machines.

@@

Unfortunately, the lens equation and the arrival times alone are not enough to form a
closed volume in the solution space. We have not imposed any strong constraints
on the density even so much as to ensure that it is non-negative everywhere. We
must therefore provide additional linear constraints so that we only sample
physical models. The explicit implementation for these priors has been
explained in \cite{2008ApJ...679...17C}, but can be summarized as
\begin{enumerate}
\item The density must be non-negative everywhere.
\item The density profile must have a slope everywhere $\le 0$.
\item The local gradient everywhere must point within $45^{\circ}$ of the center.
\item Density variations must be smooth.
\item Image parity is enforced.
\item The density is radially symmetric. (Optional)
\end{enumerate}

@@

\subsection{Removing models with extra images}\label{sec:glassextraimages} 
While linear constraints are applied in \Glass\ by the nature of the sampling
algorithm, non-linear constraints must be applied in post-processing. Models
that are inconsistent with such constraints must then be statistically
discarded via a likelihood analysis. An example of such a non-linear constraint
is the spurious presence of unobserved images. This `null-space' prior was
first proposed and explored by \citet{2006MNRAS.367.1209L} and found to be
extremely powerful. We find that our gradient prior in \Glass\ (see
\S\ref{sec:discrete}), performs much of the same function as
\citeauthor{2006MNRAS.367.1209L}'s null-space prior, but some models can still
rarely turn up spurious images. We reject these in a post-processing step,
where we sweep through the model ensemble applying the ray tracing algorithm
described in \ref{Raytracing}. \hilight{JR. Typically, this results in a rejection rate of XXX\% IS this working? Can we run it once and add the number here?}.

@@

Each of these configurations were modelled with and without time delays; with
and without a central image; and with and without the stellar mass as a lower
bound, for a total of 48 test cases. (The central image is typically highly
demagnified. For galaxy lenses it is very difficult to find since it lies along
the sight line to the bright lensing galaxy; in clusters, however, such images
have been seen -- e.g. \citealt{2005PASJ...57L...7I}). We assumed for all our tests that the
lensing mass was radially symmetric (Prior vi). For our mock data, this is
known to be true; it is most often the case with real galaxies, unless there is
an obvious observed asymmetry. The central pixel was refined into a further
$5\times5$ pixels (three extra radial positions) to capture any steep rise in
the profile.  Two of the four mock galaxies have a steeply rising inner
profile. \hilight{JR.  CAN WE DO THIS? What happens if we don't assume this symmetry? We ought
to add an appendix showing this for one example case.}

@@

We can also more directly probe the recovery of the shape of the mass
distribution by defining a global measure of lens shape
%
\begin{equation} 
    s \equiv \lambda_1/\lambda_2
\end{equation} 
%
where $\lambda_1$ and $\lambda_2$ are the eigenvalues of the 2D inertia tensor
%
\begin{equation} 
    I_{ij} = (-1)^{1-\delta_{ij}}\sum_\theta M(\vec\theta) \theta_i\theta_j 
\label{eqn:inertia}
\end{equation}
%
where $\delta_{ij}$ is the Kronecker delta. We always take $\lambda_1$ to be the largest value. As with $f_R$ and $f_\theta$, we only consider mass up to one radial position passed the outermost image and compute the fractional error as:
%
\begin{equation} \label{ferror shape}
  f_\mathrm{shape} = \left|s - \tilde s\right|  / \tilde s
\end{equation}
%
where $\tilde s$ is the shape of the mock galaxy. The distribution of
$f_\mathrm{shape}$ for each mock galaxy and each test case is shown in
\figref{shape results}.  Interestingly, for this global shape parameter
recovery it appears more important to have time delay data and/or a central
image (T,M) than to have a quad or multiple sources with wide redshift
separation. In all cases, the stellar mass little-aids the recovery, reflecting
the fact that $s$ is heavily weighted towards the shape at the {\it edge} of
the mass map, rather than at the centre where the stars may dominate the
potential (see equation \ref{eqn:inertia}).

@@

\section{No symmetric prior}
\begin{figure*}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-nosymm_Tms-kappa.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-nosymm_Tms-a.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-nosymm_Tms-b.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-nosymm_Tms-c.pdf} \\
\includegraphics[width=0.24\textwidth]{BCExtendedDoubleR1-nosymm_tms-kappa.pdf}
\includegraphics[width=0.24\textwidth]{BCExtendedDoubleR1-nosymm_tms-a.pdf}
\includegraphics[width=0.24\textwidth]{BCExtendedDoubleR1-nosymm_tms-b.pdf}
\includegraphics[width=0.24\textwidth]{BCExtendedDoubleR1-nosymm_tms-c.pdf}
\caption{No symmetry prior.}
\label{}
\end{figure*}

\section{Pixel resolution convergence test}
\begin{figure*}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr10_TmS-a.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr10_TmS-b.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr10_TmS-c.pdf} \\
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr7_TmS-a.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr7_TmS-b.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr7_TmS-c.pdf} \\
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr6_TmS-a.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr6_TmS-b.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr6_TmS-c.pdf} \\
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr5_TmS-a.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr5_TmS-b.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr5_TmS-c.pdf} \\
\caption{Changing pixrad. top to bottom: pixrad=10,7,6,5. We use 8 normally.}
\label{}
\end{figure*}

\section{Pixel resolution convergence test}
\hilight{JR. Need to add an appendix showing convergence with increasing Pixrad.}

@@

\cite{2007ApJ...667..645R} were the first to test the recovery of
strong lensing surface density profiles using mock data generated from
N-body models {\bf JR. Is this true?}. They found that when including
time delays, circularly averaged surface density distributions can
sufficiently well-recovered to make interesting comparisons with
theoretical models Cosmological parameters, however, are more
sensitive. Since then, several groups have used N-body mock data to
explore the fidelity of strong lensing reconstructions
\citep{2007MNRAS.380.1729L}. We expand on these earlier works
by considering not only the recovery of the circularly averaged
surface density profile of a lens, but also its shape, higher order
measures of the mass recovery (a pixel-by-pixel comparison), and the
utility of wrapping in constraints from stellar mass estimates and/or
stellar kinematics.

>>>>

\title[Lens Recovery with \Glass]{Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens}
\author[J.~P.~Coles~et~al.]{%
Jonathan P.~Coles,$^{1,3}$\thanks{jonathan@exascale-computing.eu}%
\newauthor%
Justin I.~Read$^2$,%
\newauthor%
and Prasenjit Saha$^3$%
%
\\
$^1$Exascale Research Computing Lab, Universit\'e de Versailles Saint-Quentin-en-Yvelines,\\$\quad$55 Avenue des Etats Unis, 78000, Versailles, France\\
$^2$Department of Physics, University of Surrey, Guildford, Surrey, GU2 7XH, United Kingdom \\
$^3$Institute of Physics, University of Zurich, 190 Winterthurerstrasse, 8057, Zurich, Switzerland
}

@@

The key scientific and technical improvements are:
%
\begin{enumerate}
  \setcounter{enumi}{0}
  \item A modular framework that allows new priors to be added and modified easily.
\end{enumerate}
%
Each prior is a simple function that adds linear constraints that operate on either
a single lens object or the entire ensemble of objects. \Glass\ comes with a number
of useful priors, the default ones will be described later, but a user can
write their own directly in the input file, or by modifying the source code.
%
\begin{enumerate}
  \setcounter{enumi}{1}
  \item The basis functions approximating a model can be changed. 
\end{enumerate}
%
\Glass\ currently describes the lens mass as a collection of pixels, but the code
has been designed to support alternative methods. In particular, there are future
plans to develop a module using Bessel functions. This will require a new set of 
priors that operate on these functions.
%
\begin{enumerate}
  \setcounter{enumi}{2}
  \item Non-linear constraints can be imposed in an automated post-processing step. 
\end{enumerate}
%
Once \Glass\ has generated an ensemble of models given the linear constraints, any number
of post processing functions can be applied. Not only can these functions be used to
derive new quantities from the mass models, they can also be used as a filter to 
accept or reject a model based on some non-linear constraint. The plotting functions
within \Glass\ will correctly display models that have been accepted or rejected.
%
\begin{enumerate}
  \setcounter{enumi}{3}
  \item The central region can have a higher resolution to capture steep models. 
\end{enumerate}
%
With the default basis set of pixels, the mass distribution of the lens is
described by a uniform grid. However, in the central region of a lensing galaxy
where the mass profile may rise steeply, the center pixel uses a higher
resolution. This allows the density to increase smoothly but still allow for
a large degree of freedom within the inner region without allowing the density
to be arbitrarily high. 
%
\begin{enumerate}
  \setcounter{enumi}{4}
  \item Stellar density can be used as an additional constraint.  
\end{enumerate}
%
The mass in inner regions of galaxies is often dominated by the stellar component
which one can estimate using standard mass-to-light models. This data can be added
to the potential as described later in \secref{stellar mass}. By using the stellar
mass one can place a lower bound on the mass and help constrain the inner most
mass profile.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item Point or extended mass objects can be placed in the field.
\end{enumerate}
%
\PixeLens\ only allows for a shear term to be added to the potential (as shown
later in \eqnref{shear}) to account for mass external to the modelled region. This
is useful to capture the gross effects of a distant neighbour. \Glass\ also
has a shear term but additionally allows for analytic potentials to
be included. This can be used to model substructure or multiple neighbours close
to the main lens. The substructure may have only a small effect if the lens is
a single galaxy, but if the lens is a group then a potential can be added for
each of the galaxies. A few standard functions are already included in \Glass\
including those for a point mass, a power law distribution, or an isothermal (a
particular case of the power law).
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item A new uniform sampling algorithm for high dimensional spaces.
\end{enumerate}
%
At the heart of \Glass\ lies a new algorithm for sampling the high dimensional
linear space that represents the modelling solution space. This algorithm was
described and tested in \cite{2012MNRAS.425.3077L}; it is 
multi-threaded allowing it to run efficiently on many-cored machines.

@@

Unfortunately, the lens equation and the arrival times alone are not enough to form a
closed volume in the solution space. We have not imposed any strong constraints
on the density even so much as to ensure that it is non-negative everywhere. We
must therefore provide additional linear constraints so that we only sample
physical models. The explicit implementation for these priors has been
explained in \cite{2008ApJ...679...17C}, but can be summarised as
\begin{enumerate}
\item The density must be non-negative everywhere.
\item The density profile must have a slope everywhere $\le 0$.
\item The local gradient everywhere must point within $45^{\circ}$ of the center.
\item Density variations must be smooth.
\item Image parity is enforced.
\item The density is radially symmetric. (Optional)
\end{enumerate}

@@

\subsection{Removing models with extra images}\label{sec:glassextraimages} 
While linear constraints are applied in \Glass\ by the nature of the sampling
algorithm, non-linear constraints must be applied in post-processing. Models
that are inconsistent with such constraints must then be statistically
discarded via a likelihood analysis. An example of such a non-linear constraint
is the spurious presence of unobserved images. This `null-space' prior was
first proposed and explored by \citet{2006MNRAS.367.1209L} and found to be
extremely powerful. We find that our gradient prior in \Glass\ (see
\S\ref{sec:discrete}), performs much of the same function as
\citeauthor{2006MNRAS.367.1209L}'s null-space prior, but some models can still
rarely turn up spurious images. We reject these in a post-processing step,
where we sweep through the model ensemble applying the ray tracing algorithm
described in \ref{Raytracing}.

@@

Each of these configurations were modelled with and without time delays; with
and without a central image; and with and without the stellar mass as a lower
bound, for a total of 48 test cases. (The central image is typically highly
demagnified. For galaxy lenses it is very difficult to find since it lies along
the sight line to the bright lensing galaxy; in clusters, however, such images
have been seen -- e.g. \citealt{2005PASJ...57L...7I}). We assumed for all our tests that the
lensing mass was radially symmetric (Prior vi). For our mock data, this is
known to be true; it is most often the case with real galaxies, unless there is
an obvious observed asymmetry. (We explore the effect of switching off the symmetry prior in Appendix \ref{no_symm_prior}. For the quads, the difference is small; for the doubles -- as expected -- the results are significantly degraded without this prior.) We use, as default, 8 pixels from the centre to the edge of the mass map; the central pixel was further refined into $5\times5$ pixels to capture any steep rise in the profile (two of the four mock galaxies have a steeply rising inner profile). We demonstrate that our results are robust to changing the grid resolution in Appendix \ref{pix_convergence_test}.

@@

We can also more directly probe the recovery of the shape of the mass
distribution by defining a global measure of lens shape
%
\begin{equation} 
    s \equiv \lambda_1/\lambda_2
\end{equation} 
%
where $\lambda_1$ and $\lambda_2$ are the eigenvalues of the 2D inertia tensor
%
\begin{equation} 
    I_{ij} = (-1)^{1-\delta_{ij}}\sum_\theta M(\vec\theta) \theta_i\theta_j 
\label{eqn:inertia}
\end{equation}
%
and $\delta_{ij}$ is the Kronecker delta. We always take $\lambda_1$ to be the largest value. As with $f_R$ and $f_\theta$, we only consider mass up to one radial position passed the outermost image and compute the fractional error as:
%
\begin{equation} \label{ferror shape}
  f_\mathrm{shape} = \left|s - \tilde s\right|  / \tilde s
\end{equation}
%
where $\tilde s$ is the shape of the mock galaxy. The distribution of
$f_\mathrm{shape}$ for each mock galaxy and each test case is shown in
\figref{shape results}.  Interestingly, for this global shape parameter
recovery it appears more important to have time delay data and/or a central
image (T,M) than to have a quad or multiple sources with wide redshift
separation. In all cases, the stellar mass little-aids the recovery, reflecting
the fact that $s$ is heavily weighted towards the shape at the {\it edge} of
the mass map, rather than at the centre where the stars may dominate the
potential (see equation \ref{eqn:inertia}).

@@

\section{No radial symmetry prior}\label{no_symm_prior}

In this appendix, we explore the effect of the radial symmetry prior. \figref{fig:no_symm_prior} shows results for a single quad (top) and an extended double (bottom); both without time delay data or stellar mass constraints. The results for the quad a similarly good to the case with the radial prior (\figref{reconstruction}); for the double, however, the results are significantly degraded as expected.

\begin{figure*}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-nosymm_Tms-kappa.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-nosymm_Tms-a.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-nosymm_Tms-b.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-nosymm_Tms-c.pdf}\\
\includegraphics[width=0.24\textwidth]{BCExtendedDoubleR1-nosymm_tms-kappa.pdf}
\includegraphics[width=0.24\textwidth]{BCExtendedDoubleR1-nosymm_tms-a.pdf}
\includegraphics[width=0.24\textwidth]{BCExtendedDoubleR1-nosymm_tms-b.pdf}
\includegraphics[width=0.24\textwidth]{BCExtendedDoubleR1-nosymm_tms-c.pdf}
\caption{Results for a single quad (top) and an extended double (bottom) without the radial symmetry prior; both without time delay data or stellar mass constraints. Figures and symbols are as in \figref{reconstruction} and \figref{2d mass reconstruction}.}
\label{fig:no_symm_prior}
\end{figure*}

\section{Pixel resolution convergence test}\label{pix_convergence_test}

In this Appendix, we present a convergence test of our results with the grid resolution -- pixrad. As default, this is set to 8 pixels from the centre of the mass map to the edge. As can be seen from \figref{fig:pix_convergence_test}, our results are typically well-converged for pixrad$\,\,> 5$. The results for pixrad$\,\,= 5$ become systematically biased away from the central regions (where we have the higher resolution adaptive mesh), because our regularisation prior combined with a low pixrad biases us towards shallow models. This effect diminishes with increasing resolution and is already negligible by pixrad$\,\,= 7$.

\begin{figure*}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr10_TmS-a.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr10_TmS-b.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr10_TmS-c.pdf} \\
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr7_TmS-a.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr7_TmS-b.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr7_TmS-c.pdf} \\
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr6_TmS-a.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr6_TmS-b.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr6_TmS-c.pdf} \\
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr5_TmS-a.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr5_TmS-b.pdf}
\includegraphics[width=0.24\textwidth]{BCQuadR1a-pr5_TmS-c.pdf} \\
\caption{Changing pixrad. top to bottom: pixrad=10,7,6,5. We use 8 normally.}
\label{fig:pix_convergence_test}
\end{figure*}

@@

%\cite{2007ApJ...667..645R} were the first to test the recovery of
%strong lensing surface density profiles using mock data generated from
%N-body models {\bf JR. Is this true?}. They found that when including
%time delays, circularly averaged surface density distributions can
%sufficiently well-recovered to make interesting comparisons with
%theoretical models Cosmological parameters, however, are more
%sensitive. Since then, several groups have used N-body mock data to
%explore the fidelity of strong lensing reconstructions
%\citep{2007MNRAS.380.1729L}. We expand on these earlier works
%by considering not only the recovery of the circularly averaged
%surface density profile of a lens, but also its shape, higher order
%measures of the mass recovery (a pixel-by-pixel comparison), and the
%utility of wrapping in constraints from stellar mass estimates and/or
%stellar kinematics.
%
